---
title: "Check "
author: "DS hearing group"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```

## Load R libraries
```{r load_libraries, warning = FALSE, message = FALSE}
library(data.table)
library(ggplot2)
library(ggpubr)
library(stringr)
library(readxl)
library(tidyr)

theme_set(theme_classic())
```

# Demographic data

## Demographic data from US Census

```{r fig.width=6, fig.height=4}
years = 2011:2021
dat_all = NULL
zips  = 98004:98008

age_grp1 = paste0("Total!!Estimate!!AGE!!", 
                 c("Under 5 years", "5 to 9 years", "10 to 14 years"))

age_grp2 = paste0("Estimate!!Total!!Total population!!AGE!!", 
                 c("Under 5 years", "5 to 9 years", "10 to 14 years"))

for(y1 in years){
  dat = fread(sprintf("US_census/ACSST5Y%d.S0101-Data.csv", y1), skip=1)
  dim(dat)
  dat[1:5,1:3]

  
  if(y1 <= 2016){
    cols2use = c("Geographic Area Name", 
                 "Total!!Estimate!!Total population", 
                 age_grp1)
  }else{
    cols2use = c("Geographic Area Name", 
                 "Estimate!!Total!!Total population", 
                 age_grp2)
  }
  
  stopifnot(all(cols2use %in% names(dat)))

  dat = dat[,..cols2use]
  names(dat) = c("zipcode", "total", "0-4", "5-9", "10-14")
  dat$zipcode = gsub("ZCTA5 ", "", dat$zipcode)
  
  if(y1 <= 2016){
    dat[,3:5] = round(dat[,3:5]*dat$total/100)
  }
  
  dat = rbind(dat, data.table(zipcode="all", t(colSums(dat[,-1]))))
  print(y1)
  print(dat)
  dat$Year = y1

  dat_all = rbind(dat_all, dat)
}

dim(dat_all)
dat_all[1:10,]

df1 = pivot_longer(dat_all, cols=c("0-4", "5-9", "10-14"),
                   names_to = "Age Group", values_to = "Total")
df1 = as.data.frame(df1)
dim(df1)
df1[1:3,]

df1$`Age Group` = factor(df1$`Age Group`, 
                         levels=c("0-4", "5-9", "10-14"))

for(zip1 in c(98004:98008, "all")){
  if(zip1 == "all"){gtitle = "98004-98008"}else{gtitle=zip1}
  df2 = df1[df1$zipcode==zip1,]
  p1 = ggplot(df2, aes(x=Year, y=Total, 
                          color=`Age Group`, shape=`Age Group`)) +
    geom_point() + geom_smooth(method=loess) + 
    scale_x_continuous("Year", seq(2011, 2021, by=2), 
                       seq(2011, 2021, by=2)) + 
    ggtitle(zip1)
  
  print(p1)
}

df1[df1$zipcode == "all",]
```


## Demographic data from censusreporter

> Census Reporter is an independent project to make data from the American Community Survey (ACS) easier to use. We are unaffiliated with the U.S. Census Bureau.

* B01001003: "Under 5 years"
* B01001004: "5 to 9 years"
* B01001005: "10 to 14 years"

We comapre 1 year vs. 5 year estimate with confidence intervals. Here are the official statemet of the two estimates

https://www.census.gov/programs-surveys/acs/guidance/estimates.html

1 year estimate: Best used when currency is more important than precision Analyzing large populations

5 year estimate: Best used when precision is more important than currency, analyzing very small populations, examining tracts and other smaller geographies because 1-year estimates are not available

> The MOEs for published ACS estimates are provided at a 90 percent
confidence level.

```{r fig.width=6, fig.height=3}

dir_5y  = "acs2021_5yr_B01001_15000US530330228033"
file_5y = "acs2021_5yr_B01001_15000US530330228033.csv"

dat_5y = fread(file.path("censusreporter", dir_5y, file_5y))
dim(dat_5y)
dat_5y[c(1:2, nrow(dat_5y)),1:12]

cols_total = c("B01001003", "B01001004", "B01001005")
cols_error = paste0(cols_total, ", Error")

nr = nrow(dat_5y)

bsd_5y = data.frame(age = c("0-4", "5-9", "10-14"), 
                    total = unlist(dat_5y[nr, ..cols_total]), 
                    error = unlist(dat_5y[nr, ..cols_error]))

dir_1y  = "acs2021_1yr_B01001_97000US5300390"
file_1y = "acs2021_1yr_B01001_97000US5300390.csv"

dat_1y = fread(file.path("censusreporter", dir_1y, file_1y))
dim(dat_1y)
dat_1y[,1:12]

bsd_1y = data.frame(age = c("0-4", "5-9", "10-14"), 
                    total = unlist(dat_1y[1, ..cols_total]), 
                    error = unlist(dat_1y[1, ..cols_error]))
bsd_1y$method = "US Census 1 year estimate"
bsd_5y$method = "US Census 5 year estimate"

bsd = rbind(bsd_1y, bsd_5y)
bsd$age = factor(bsd$age, levels = c("0-4", "5-9", "10-14"))

ci_ratio = qnorm(0.975)/qnorm(0.95)
ci_ratio

ggplot(bsd, aes(x = age, y = total, fill=method)) +
  geom_bar(stat = "identity", position=position_dodge()) +
  geom_errorbar(aes(ymin = total - ci_ratio*error, 
                    ymax = total + ci_ratio*error),
                width = 0.2, position = position_dodge(0.9)) +
  labs(title = "BSD: US census total and 95% confidence interval")

bsd = data.table(bsd)
bsd[,CI:=paste0(round(total - ci_ratio*error), "-", 
                round(total + ci_ratio*error))]

bsd
```

# Enrollment data

```{r}
fnm = "Report_Card_Enrollment_from_2014-15_to_Current_Year.csv"
dat = fread(paste0("Report_Card_Enrollment/", fnm))

dim(dat)
dat[1:2,]

table(dat$SchoolYear)
table(dat$OrganizationLevel)

table(dat$DistrictName == "Bellevue School District")

bsd = dat[DistrictName == "Bellevue School District",]
dim(bsd)
bsd[1:2,]

table(bsd$SchoolName)

bsd_total = bsd[SchoolName == "District Total",]
somerset  = bsd[SchoolName == "Somerset Elementary School",]

dim(bsd_total)
dim(somerset)

bsd_total[1:2,]

table(bsd_total$SchoolYear)
table(somerset$SchoolYear)

bsd_total[SchoolYear=="2014-15", Gradelevel]
bsd_total[SchoolYear=="2020-21", Gradelevel]

somerset[SchoolYear=="2014-15", Gradelevel]
somerset[SchoolYear=="2021-22", Gradelevel]

grads = somerset[SchoolYear=="2021-22", Gradelevel]
grads = setdiff(grads, "AllGrades")

grads

fnm = "Report_Card_Enrollment_2022-23_School_Year.csv"
dat_23 = fread(paste0("Report_Card_Enrollment/", fnm))
dim(dat_23)
dat_23[1:2,]

names(dat_23)[names(dat_23)=="schoolyear"] = "SchoolYear"
names(dat)[names(dat)=="Gradelevel"] = "GradeLevel"

cols = c("SchoolYear", "DistrictName", "SchoolName", 
         "GradeLevel", "All Students")
table(cols %in% names(dat))
table(cols %in% names(dat_23))

dat = rbind(dat[,..cols], dat_23[,..cols])
dim(dat)
dat[1:2,]
```

## Summarize one school district

```{r fig.width=8, fig.height=3}

plot_district <- function(dat, school_district, grads){
  
  grads_grp1 = grads[c(1,2,6)]
  grads_grp2 = grads[3:5]

  district = dat[DistrictName == school_district]
  district = district[SchoolName == "District Total",]
  district = district[GradeLevel %in% grads,]
  district$SchoolYear = gsub("^20", "", district$SchoolYear)
  district$`All Students` = as.numeric(gsub(",", "", 
                                            district$`All Students`))
  
  p1 = ggplot(subset(district, GradeLevel %in% grads_grp1), 
              aes(x=SchoolYear, y=`All Students`, 
                          color=GradeLevel, group=GradeLevel)) +
    geom_line(color="grey") + geom_point() + 
    ylim(0, max(district$`All Students`)) + 
    ggtitle(school_district) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  p2 = ggplot(subset(district, GradeLevel %in% grads_grp2), 
              aes(x=SchoolYear, y=`All Students`, 
                          color=GradeLevel, group=GradeLevel)) +
    geom_line(color="grey") + geom_point() +  
    ylim(0, max(district$`All Students`)) + 
    ggtitle(school_district) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggarrange(p1, p2, nrow=1, ncol=2)
}

plot_district(dat, "Bellevue School District", grads)
plot_district(dat, "Seattle School District No. 1", grads)
plot_district(dat, "Lake Washington School District", grads)
plot_district(dat, "Issaquah School District", grads)

```

## Summarize one school

```{r fig.width=8, fig.height=3}

plot_school <- function(dat, school, grads){
  
  grads_grp1 = grads[c(1,2,6)]
  grads_grp2 = grads[3:5]

  df = dat[SchoolName == school & DistrictName=="Bellevue School District",]
  df = df[GradeLevel %in% grads,]
  df$SchoolYear = gsub("^20", "", df$SchoolYear)
  df$`All Students` = as.numeric(gsub(",", "", 
                                      df$`All Students`))
  
  p1 = ggplot(subset(df, GradeLevel %in% grads_grp1), 
              aes(x=SchoolYear, y=`All Students`, 
                          color=GradeLevel, group=GradeLevel)) +
    geom_line(color="grey") + geom_point() +  
    ylim(0, max(df$`All Students`)) + 
    ggtitle(school) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  p2 = ggplot(subset(df, GradeLevel %in% grads_grp2), 
              aes(x=SchoolYear, y=`All Students`, 
                          color=GradeLevel, group=GradeLevel)) +
    geom_line(color="grey") + geom_point() +  
    ylim(0, max(df$`All Students`)) + 
    ggtitle(school) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggarrange(p1, p2, nrow=1, ncol=2)
}

table(dat$SchoolName[dat$DistrictName == "Bellevue School District" & 
                     dat$GradeLevel == "1st Grade"])
plot_school(dat, "Somerset Elementary School", grads)
plot_school(dat, "Eastgate Elementary School", grads)
plot_school(dat, "Jing Mei Elementary School", grads)
plot_school(dat, "Medina Elementary School", grads)
plot_school(dat, "Wilburton Elementary School", grads)
plot_school(dat, "Cherry Crest Elementary School", grads)

```

# Appendix

## Demographic data from Washington state OFM

Small Area Demographic Estimates (SADE) by School district, downloaded from Office of Financial Management (OFM) [website](https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/estimates-april-1-population-age-sex-race-and-hispanic-origin). 

Note that this data may not be accurate since as mentioned in the OFM website: 

> OFM 2011 through 2020 estimates are based on 2010 census data. Results from the 2020 Census are not available yet. OFM Small Area Demographic Estimates from 2011 forward will be revised after detailed 2020 Census data are released.


```{r fig.width=8, fig.height=3}
sade = read_excel("ofm_pop_sade_sduni_2010_to_2020.xlsx", 
                         sheet = "Total")
dim(sade)
sade[1:2,]

school_districts = c("Bellevue School District", 
                    "Lake Washington School District", 
                    "Issaquah School District",
                    "Seattle School District")

stopifnot(all(school_districts %in% sade$`Area Name`))

sade = sade[sade$`Area Name` %in% school_districts,]
table(sade$`Age Group`)

sade = as.data.table(sade[,1:7])

age_groups = c("0-4", "5-9", "10-14", "30-34", "35-39", "40-44", "45-49")
sade = sade[`Age Group` %in% age_groups, ]

table(sade$`Area Name`)
table(sade$Year)
sade[1:2,]
sade$Year   = as.numeric(sade$Year)
sade$Total  = as.numeric(sade$Total)
sade$Male   = as.numeric(sade$Male)
sade$Female = as.numeric(sade$Female)

sade$Age_set = "Young_adults"
sade[`Age Group` %in% c("0-4", "5-9", "10-14"), Age_set:= "kids"]

dim(sade)
sade[1:2,]

sade$`Age Group` = factor(sade$`Age Group`, 
                                 levels=age_groups)
```

### Bellevue
```{r fig.width=8, fig.height=3}

plot1 <- function(df, age, area){
  df1 = df[df$Age_set == age & df$`Area Name` == area, ]
  g1 = ggplot(df1, aes(x=Year, y=Total, color=`Age Group`, shape=`Age Group`)) +
    geom_point() + geom_smooth(method=loess) + 
    scale_x_continuous("Year", seq(2010, 2020, by=2), seq(2010, 2020, by=2)) + 
    ggtitle(area)
  g1
}

g1 = plot1(sade, age="kids", area="Bellevue School District")
g2 = plot1(sade, age="Young_adults", area="Bellevue School District")

ggarrange(g1, g2)
```

### Seattle

```{r fig.width=8, fig.height=3}

g1 = plot1(sade, age="kids", area="Seattle School District")
g2 = plot1(sade, age="Young_adults", area="Seattle School District")

ggarrange(g1, g2)
```


### Lake Washington

```{r fig.width=8, fig.height=3}

g1 = plot1(sade, age="kids", area="Lake Washington School District")
g2 = plot1(sade, age="Young_adults", area="Lake Washington School District")

ggarrange(g1, g2)
```

### Issaquah

```{r fig.width=8, fig.height=3}

g1 = plot1(sade, age="kids", area="Issaquah School District")
g2 = plot1(sade, age="Young_adults", area="Issaquah School District")

ggarrange(g1, g2)
```

# Session information
```{r}
gc()

sessionInfo()
```



